Reanalysis Step 2b: Multiple Imputation Outcome Diagnostics

### 2025-08-29 reanalysis-2b-imputation_diagnostics.qmd
enhance

# Create indicator for which outcomes were imputed
df_original <- df_original |>
  mutate(
    outcome_was_imputed = cohort_id == "excluded",
    had_intermediate = !is.na(intermediate_VA_logMAR)
  )

# Summary of imputation status
imputation_summary <- df_original |>
  group_by(outcome_was_imputed) |>
  summarise(
    n = n(),
    pct = round(100 * n / nrow(df_original), 1),
    had_baseline = sum(!is.na(baseline_VA_logMAR)),
    had_intermediate = sum(had_intermediate),
    .groups = "drop"
  )

kable(imputation_summary,
      col.names = c("Outcome Imputed", "N", "%", "Has Baseline", "Has Intermediate"),
      caption = "Summary of imputation status for outcome VA")
```

@@

```{r calculate-changes}
# Function to calculate change scores for a single imputed dataset
calculate_change_scores <- function(imp_data, original_data) {
  imp_data |>
    mutate(
      # Add imputation status from original data
      outcome_was_imputed = original_data$outcome_was_imputed,
      had_intermediate = original_data$had_intermediate,

@@

```{r summary-stats}
# Compare change scores between imputed and observed
change_summary <- changes_df |>
  group_by(outcome_was_imputed) |>
  summarise(
    n = n(),
    mean_change = mean(change_baseline_to_outcome, na.rm = TRUE),
    sd_change = sd(change_baseline_to_outcome, na.rm = TRUE),
    median_change = median(change_baseline_to_outcome, na.rm = TRUE),
    q25_change = quantile(change_baseline_to_outcome, 0.25, na.rm = TRUE),
    q75_change = quantile(change_baseline_to_outcome, 0.75, na.rm = TRUE),
    min_change = min(change_baseline_to_outcome, na.rm = TRUE),
    max_change = max(change_baseline_to_outcome, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    iqr_change = q75_change - q25_change,
    group = ifelse(outcome_was_imputed, "Imputed", "Observed")
  )

kable(change_summary |> 
        select(group, n, mean_change, sd_change, median_change, iqr_change, min_change, max_change),
      digits = 3,
      col.names = c("Group", "N", "Mean", "SD", "Median", "IQR", "Min", "Max"),
      caption = "Distribution of vision changes (outcome - baseline) in logMAR units")

# Statistical test for difference in distributions
t_test <- t.test(change_baseline_to_outcome ~ outcome_was_imputed, data = changes_df)
ks_test <- ks.test(
  changes_df$change_baseline_to_outcome[changes_df$outcome_was_imputed],
  changes_df$change_baseline_to_outcome[!changes_df$outcome_was_imputed]
)

test_results <- data.frame(
  Test = c("Welch's t-test", "Kolmogorov-Smirnov"),
  Statistic = c(t_test$statistic, ks_test$statistic),
  P_Value = c(t_test$p.value, ks_test$p.value),
  Interpretation = c(
    ifelse(t_test$p.value < 0.05, "Means differ significantly", "Means similar"),
    ifelse(ks_test$p.value < 0.05, "Distributions differ significantly", "Distributions similar")
  )
)

kable(test_results, digits = 4, caption = "Statistical tests comparing change distributions")
```

@@

# Prepare data for plotting
plot_data <- changes_df |>
  mutate(
    Group = ifelse(outcome_was_imputed, "Imputed", "Observed"),
    Group = factor(Group, levels = c("Observed", "Imputed"))
  ) |>
  filter(!is.na(change_baseline_to_outcome))

# 1. Density plots
p1 <- ggplot(plot_data, aes(x = change_baseline_to_outcome, fill = Group)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_vline(data = change_summary, 
             aes(xintercept = mean_change, color = group),
             linetype = "solid", size = 1) +
  scale_fill_manual(values = c("Observed" = "#2E86AB", "Imputed" = "#A23B72")) +
  scale_color_manual(values = c("Observed" = "#2E86AB", "Imputed" = "#A23B72")) +
  labs(
    title = "Distribution of Vision Changes",
    subtitle = "Vertical lines show group means",
    x = "Change in logMAR (positive = worsening)",
    y = "Density"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

# 2. Boxplots
p2 <- ggplot(plot_data, aes(x = Group, y = change_baseline_to_outcome, fill = Group)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  scale_fill_manual(values = c("Observed" = "#2E86AB", "Imputed" = "#A23B72")) +
  labs(
    title = "Vision Changes by Group",
    subtitle = "Box shows IQR, whiskers extend to 1.5*IQR",
    y = "Change in logMAR (positive = worsening)",
    x = ""
  ) +
  theme_minimal() +
  theme(legend.position = "none")

@@

# Arrange plots
grid.arrange(p1, p2, p3, p4, ncol = 2)
```

@@

```{r directional-changes}
# Compare proportions of improvement/stable/worsening
direction_summary <- changes_df |>
  filter(!is.na(outcome_direction)) |>
  group_by(outcome_was_imputed, outcome_direction) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(outcome_was_imputed) |>
  mutate(
    total = sum(n),
    pct = round(100 * n / total, 1),
    Group = ifelse(outcome_was_imputed, "Imputed", "Observed")
  )

@@

>>>>

# Create more nuanced grouping
df_original <- df_original |>
  mutate(
    outcome_was_imputed = cohort_id == "excluded",
    had_intermediate = !is.na(intermediate_VA_logMAR),

    # Create detailed imputation groups
    imputation_group = case_when(
      !outcome_was_imputed ~ "Observed",
      outcome_was_imputed & had_intermediate ~ "Imputed (with intermediate)",
      outcome_was_imputed & !had_intermediate ~ "Imputed (no intermediate)",
      TRUE ~ "Unknown"
    ),
    imputation_group = factor(imputation_group, 
                             levels = c("Observed", 
                                       "Imputed (with intermediate)", 
                                       "Imputed (no intermediate)"))
  )

# Summary of imputation status
imputation_summary <- df_original |>
  group_by(imputation_group) |>
  summarise(
    n = n(),
    pct = round(100 * n / nrow(df_original), 1),
    had_baseline = sum(!is.na(baseline_VA_logMAR)),
    pct_baseline = round(100 * had_baseline / n, 1),
    .groups = "drop"
  )

kable(imputation_summary,
      col.names = c("Group", "N", "%", "Has Baseline", "% Baseline"),
      caption = "Detailed breakdown of imputation groups")

# Baseline characteristics by group
baseline_chars <- df_original |>
  group_by(imputation_group) |>
  summarise(
    # Demographics
    mean_age = mean(age, na.rm = TRUE),
    sd_age = sd(age, na.rm = TRUE),
    pct_male = round(100 * mean(gender == "Male", na.rm = TRUE), 1),

    # Clinical characteristics
    mean_CCI = mean(CCI, na.rm = TRUE),
    sd_CCI = sd(CCI, na.rm = TRUE),
    mean_DCSI = mean(DCSI, na.rm = TRUE),
    sd_DCSI = sd(DCSI, na.rm = TRUE),

    # Baseline vision
    mean_baseline_VA = mean(baseline_VA_logMAR, na.rm = TRUE),
    sd_baseline_VA = sd(baseline_VA_logMAR, na.rm = TRUE),

    # Treatment indicators
    pct_anti_VEGF = round(100 * mean(anti_VEGF == 1, na.rm = TRUE), 1),
    pct_PRP = round(100 * mean(PRP_flag == 1, na.rm = TRUE), 1),

    # Lapse status
    pct_lapsed = round(100 * mean(ever_lapse_binary == 1, na.rm = TRUE), 1),

    .groups = "drop"
  )

kable(baseline_chars |> 
        mutate(across(where(is.numeric), ~round(., 2))),
      caption = "Baseline characteristics by imputation group")
```

@@

```{r calculate-changes}
# Function to calculate change scores for a single imputed dataset
calculate_change_scores <- function(imp_data, original_data) {
  imp_data |>
    mutate(
      # Add imputation status from original data
      outcome_was_imputed = original_data$outcome_was_imputed,
      had_intermediate = original_data$had_intermediate,
      imputation_group = original_data$imputation_group,

@@

```{r summary-stats}
# Compare change scores across all three groups
change_summary_detailed <- changes_df |>
  group_by(imputation_group) |>
  summarise(
    n = n(),
    mean_change = mean(change_baseline_to_outcome, na.rm = TRUE),
    sd_change = sd(change_baseline_to_outcome, na.rm = TRUE),
    median_change = median(change_baseline_to_outcome, na.rm = TRUE),
    q25_change = quantile(change_baseline_to_outcome, 0.25, na.rm = TRUE),
    q75_change = quantile(change_baseline_to_outcome, 0.75, na.rm = TRUE),
    min_change = min(change_baseline_to_outcome, na.rm = TRUE),
    max_change = max(change_baseline_to_outcome, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    iqr_change = q75_change - q25_change
  )

kable(change_summary_detailed |> 
        select(imputation_group, n, mean_change, sd_change, median_change, iqr_change, min_change, max_change),
      digits = 3,
      col.names = c("Group", "N", "Mean", "SD", "Median", "IQR", "Min", "Max"),
      caption = "Distribution of vision changes (outcome - baseline) by detailed imputation group")

# Also keep simplified version for compatibility
change_summary <- changes_df |>
  group_by(outcome_was_imputed) |>
  summarise(
    n = n(),
    mean_change = mean(change_baseline_to_outcome, na.rm = TRUE),
    sd_change = sd(change_baseline_to_outcome, na.rm = TRUE),
    median_change = median(change_baseline_to_outcome, na.rm = TRUE),
    q25_change = quantile(change_baseline_to_outcome, 0.25, na.rm = TRUE),
    q75_change = quantile(change_baseline_to_outcome, 0.75, na.rm = TRUE),
    min_change = min(change_baseline_to_outcome, na.rm = TRUE),
    max_change = max(change_baseline_to_outcome, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    iqr_change = q75_change - q25_change,
    group = ifelse(outcome_was_imputed, "Imputed", "Observed")
  )

# Statistical tests across three groups
anova_test <- aov(change_baseline_to_outcome ~ imputation_group, data = changes_df)
anova_summary <- summary(anova_test)

# Pairwise comparisons
pairwise_tests <- pairwise.t.test(changes_df$change_baseline_to_outcome, 
                                  changes_df$imputation_group,
                                  p.adjust.method = "bonferroni")

# Format ANOVA results
anova_results <- data.frame(
  Test = "One-way ANOVA",
  F_Statistic = anova_summary[[1]]$`F value`[1],
  df1 = anova_summary[[1]]$Df[1],
  df2 = anova_summary[[1]]$Df[2],
  P_Value = anova_summary[[1]]$`Pr(>F)`[1],
  Interpretation = ifelse(anova_summary[[1]]$`Pr(>F)`[1] < 0.05, 
                         "Groups differ significantly", "No significant difference")
)

kable(anova_results, digits = 4, caption = "ANOVA comparing change distributions across three groups")

# Format pairwise results
pairwise_matrix <- pairwise_tests$p.value
pairwise_df <- data.frame(
  Comparison = c("Imputed (with intermediate) vs Observed",
                "Imputed (no intermediate) vs Observed",
                "Imputed (no intermediate) vs Imputed (with intermediate)"),
  P_Value = c(pairwise_matrix[1,1], pairwise_matrix[2,1], pairwise_matrix[2,2]),
  Significant = ifelse(c(pairwise_matrix[1,1], pairwise_matrix[2,1], pairwise_matrix[2,2]) < 0.05,
                      "Yes", "No")
)

kable(pairwise_df, digits = 4, caption = "Pairwise comparisons (Bonferroni adjusted)")

# Also do simplified tests for backward compatibility
t_test <- t.test(change_baseline_to_outcome ~ outcome_was_imputed, data = changes_df)
ks_test <- ks.test(
  changes_df$change_baseline_to_outcome[changes_df$outcome_was_imputed],
  changes_df$change_baseline_to_outcome[!changes_df$outcome_was_imputed]
)
```

@@

# Prepare data for plotting - using detailed groups
plot_data_detailed <- changes_df |>
  filter(!is.na(change_baseline_to_outcome))

# Also prepare simplified version for some plots
plot_data <- changes_df |>
  mutate(
    Group = ifelse(outcome_was_imputed, "Imputed", "Observed"),
    Group = factor(Group, levels = c("Observed", "Imputed"))
  ) |>
  filter(!is.na(change_baseline_to_outcome))

# 1. Density plots - detailed groups
p1_detailed <- ggplot(plot_data_detailed, aes(x = change_baseline_to_outcome, fill = imputation_group)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_vline(data = change_summary_detailed, 
             aes(xintercept = mean_change, color = imputation_group),
             linetype = "solid", size = 1) +
  scale_fill_manual(values = c("Observed" = "#2E86AB", 
                              "Imputed (with intermediate)" = "#A23B72",
                              "Imputed (no intermediate)" = "#F18F01")) +
  scale_color_manual(values = c("Observed" = "#2E86AB", 
                               "Imputed (with intermediate)" = "#A23B72",
                               "Imputed (no intermediate)" = "#F18F01")) +
  labs(
    title = "Distribution of Vision Changes by Detailed Group",
    subtitle = "Vertical lines show group means",
    x = "Change in logMAR (positive = worsening)",
    y = "Density"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

# 1b. Simplified density plot (for comparison)
p1 <- ggplot(plot_data, aes(x = change_baseline_to_outcome, fill = Group)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_vline(data = change_summary, 
             aes(xintercept = mean_change, color = group),
             linetype = "solid", size = 1) +
  scale_fill_manual(values = c("Observed" = "#2E86AB", "Imputed" = "#A23B72")) +
  scale_color_manual(values = c("Observed" = "#2E86AB", "Imputed" = "#A23B72")) +
  labs(
    title = "Distribution of Vision Changes (Simplified)",
    subtitle = "Vertical lines show group means",
    x = "Change in logMAR (positive = worsening)",
    y = "Density"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

# 2. Boxplots - detailed groups
p2_detailed <- ggplot(plot_data_detailed, aes(x = imputation_group, y = change_baseline_to_outcome, fill = imputation_group)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  scale_fill_manual(values = c("Observed" = "#2E86AB", 
                              "Imputed (with intermediate)" = "#A23B72",
                              "Imputed (no intermediate)" = "#F18F01")) +
  labs(
    title = "Vision Changes by Detailed Group",
    subtitle = "Box shows IQR, whiskers extend to 1.5*IQR",
    y = "Change in logMAR (positive = worsening)",
    x = ""
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

@@

# Arrange plots - show detailed version first
grid.arrange(p1_detailed, p2_detailed, ncol = 2)

# Then show additional comparisons
grid.arrange(p3, p4, ncol = 2)
```

@@

```{r directional-changes}
# Compare proportions across detailed groups
direction_summary_detailed <- changes_df |>
  filter(!is.na(outcome_direction)) |>
  group_by(imputation_group, outcome_direction) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(imputation_group) |>
  mutate(
    total = sum(n),
    pct = round(100 * n / total, 1)
  )

# Reshape for display
direction_wide_detailed <- direction_summary_detailed |>
  select(imputation_group, outcome_direction, n, pct) |>
  pivot_wider(
    names_from = outcome_direction,
    values_from = c(n, pct),
    values_fill = list(n = 0, pct = 0)
  )

kable(direction_wide_detailed,
      caption = "Direction of vision change by detailed imputation group")

# Also create simplified version for backward compatibility
direction_summary <- changes_df |>
  filter(!is.na(outcome_direction)) |>
  group_by(outcome_was_imputed, outcome_direction) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(outcome_was_imputed) |>
  mutate(
    total = sum(n),
    pct = round(100 * n / total, 1),
    Group = ifelse(outcome_was_imputed, "Imputed", "Observed")
  )

@@

## Baseline Predictors of Change

This section examines whether baseline characteristics (like CCI) predict the magnitude of change, which might explain differences between groups.

```{r baseline-predictors}
# Merge baseline characteristics with change scores
analysis_df <- changes_df |>
  left_join(df_original |> select(CCI, DCSI, age, baseline_VA_logMAR), 
            by = c("baseline_VA_logMAR"))

# Compare baseline predictors by imputation group
predictor_summary <- analysis_df |>
  group_by(imputation_group) |>
  summarise(
    n = n(),
    # Correlation between CCI and change
    cor_CCI_change = cor(CCI, change_baseline_to_outcome, use = "complete.obs"),
    # Correlation between baseline VA and change  
    cor_baseline_change = cor(baseline_VA_logMAR, change_baseline_to_outcome, use = "complete.obs"),
    # Mean change by CCI tercile
    mean_change_low_CCI = mean(change_baseline_to_outcome[CCI <= quantile(CCI, 0.33, na.rm = TRUE)], na.rm = TRUE),
    mean_change_high_CCI = mean(change_baseline_to_outcome[CCI >= quantile(CCI, 0.67, na.rm = TRUE)], na.rm = TRUE),
    .groups = "drop"
  )

kable(predictor_summary |> mutate(across(where(is.numeric), ~round(., 3))),
      caption = "Relationship between baseline characteristics and vision change")

# Regression model to check if group differences persist after adjustment
adj_model <- lm(change_baseline_to_outcome ~ imputation_group + CCI + DCSI + age + baseline_VA_logMAR,
                data = analysis_df)

adj_summary <- summary(adj_model)
adj_coef <- data.frame(
  Variable = rownames(adj_summary$coefficients),
  Estimate = adj_summary$coefficients[,1],
  SE = adj_summary$coefficients[,2],
  P_Value = adj_summary$coefficients[,4]
) |>
  filter(grepl("imputation_group", Variable))

kable(adj_coef |> mutate(across(where(is.numeric), ~round(., 4))),
      caption = "Group differences after adjusting for baseline characteristics")

# Visualization: Change by CCI tertile and group
analysis_df <- analysis_df |>
  mutate(
    CCI_tertile = cut(CCI, 
                     breaks = quantile(CCI, c(0, 0.33, 0.67, 1), na.rm = TRUE),
                     labels = c("Low", "Medium", "High"),
                     include.lowest = TRUE)
  )

p_cci <- ggplot(analysis_df |> filter(!is.na(CCI_tertile)), 
                aes(x = CCI_tertile, y = change_baseline_to_outcome, fill = imputation_group)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  facet_wrap(~ imputation_group, scales = "free_x") +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  scale_fill_manual(values = c("Observed" = "#2E86AB", 
                              "Imputed (with intermediate)" = "#A23B72",
                              "Imputed (no intermediate)" = "#F18F01")) +
  labs(
    title = "Vision Change by CCI Tertile and Imputation Group",
    x = "CCI Tertile",
    y = "Change in logMAR"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

print(p_cci)
```
